Teaching Low-Level Concurrency with \pebbles and \landslide

### 2016-08-27 sigcse-17/landslide.tex
sigcse: garth's revisions

\section{\landslide}
% TODO camready
%\section{Landslide}
\label{sec:landslide}

@@

Conceptually,
\landslide is designed as follows.
Through several annotations in the kernel source, \landslide tracks
%the scheduler state to know
which threads are running or runnable.
(For testing thread libraries, these annotations are already provided in the reference kernel binary;
for testing kernels, the student must add them by hand.)
When execution reaches certain interesting {\em preemption points},
\landslide triggers artificial clock interrupts
to force the scheduler to run a different thread.
When a test finishes execution according
to one pattern of thread switches,
\landslide rewinds the simulation state
and resumes the test according to a different interleaving.
After each instruction,
\landslide applies several bug-detection predicates to
the program state to detect
illegal heap accesses,
deadlock, infinite loops, and panics.

@@

{\bf State space exploration.}
To trigger the interleaving which exposes this bug, threads must be preempted at the boundaries of the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
\landslide automatically considers these boundaries to be {\em preemption points}: candidate sites for timer-driven thread switches.
It can also add preemption points dynamically,
as identified by an integrated data-race analysis \cite{hybriddatarace,fasttrack},
to expose shared-memory bugs when synchronization is absent.

The branching possibilities arising from each preemption point produce an execution tree, or state space,
as shown in Figure \ref{fig:paradise}(b).
%As these state spaces are exponentially-sized,
\landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
to mitigate the exponential explosion of these state spaces.
% TODO make sure this gets mentioned later, if not here
%and state space estimation \cite{estimation} to predict which
% TODO CAMREADY: cite oopsla
% In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.
This reduction requires only tracing the memory accesses of each thread,
so we are able to perform it behind-the-scenes, without consuming any student attention.

Apart from which test program to run, \landslide's user interface requires just one option:
students specify how much time they are willing to wait.
%Considering this CPU budget,
\landslide then uses state space estimation \cite{estimation} to predict
% TODO CAMREADY: deanonymize
which combinations of preemption points will produce appropriately-sized state spaces for the given CPU-time budget \cite{quicksand-anonymized}.
If no bug is found before time runs out, \landslide displays a progress report summarizing the estimated remaining time for each state space,
so ambitious students can decide whether to increase the time limit and retry.

@@

When \landslide identifies a bug, it outputs an HTML-formatted
{\em preemption trace}.
Figure~\ref{fig:trace} shows a typical output for the example bug in Figure \ref{fig:paradise}.
The trace reports what kind of bug was detected
and each preemption point in the current interleaving:
which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
Hence, students can bypass tediously reconstructing the schedule using \x{printf},
and hopefully better understand the concurrent execution that exposed the bug,

@@

For the thread library project, \landslide instruments student implementations automatically.
We provide a custom reference kernel binary equipped with the kernel instrumentations (described below)
so \landslide can track threads' lifecycles. %the lifecycle of threads.
The userspace \x{malloc} and \x{assert} are also provided in the basecode, and tracked similarly.
\landslide's data-race analysis must understand the locking discipline,
so it searches the student's code for functions named \x{mutex_lock} and \x{mutex_unlock} (as mandated by the assignment handout),
and assumes they at least provide mutual exclusion\footnote{
	To relax this assumption, we provide a special test case for mutexes,
	in which \landslide may preempt and search for data races in the mutex implementation itself.
}.
All other thread library components are treated as unknown, and subject to preemption anywhere.

To test \pebbles kernels directly, they must be annotated to inform \landslide of certain important concurrency events.
Because the kernel project leaves internal code structure as an open-ended design decision,
\landslide cannot search for specific function names. % like it does for P2s.
Instead, we provide a set of functions named with the prefix
% TODO camready
\texttt{tell\_\landslide}
%\x{tell_landslide}
to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
Users must also list all core scheduler, context switch, and timer handling routines, so \landslide can ignore their memory accesses for the sake of DPOR
(otherwise, threads would always conflict with each other and DPOR would achieve no reduction).

@@

Student submissions may behave in extraordinary ways which violate the assumptions of many modern model checkers.
\landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
but it must support testing any legal, yet distasteful, solution.
%Hence, we added several heuristic
For example, students may use open-coded \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
To avoid getting stuck in a cyclic (infinite) state space, \landslide keeps a ``yield loop'' counter for each thread;
when a thread \x{yield}s 10 times without otherwise making progress, \landslide marks it blocked.
%
In extreme cases, this may erroneously seem like deadlock when all threads \x{yield} many times coincidentally.
To avoid wasting precious student attention with false alarms, % the ultimate sin
\landslide will reset all blocked threads' yield-loop counters,
retrying potentially up to 128 times before reporting a true deadlock.

Students may also break abstraction by open-coding a \x{mutex_unlock} sequence.
\landslide must track lock operations for accurate data-race analysis,
so to catch these patterns, it
%learns the size of the \x{mutex} struct, then
checks accesses against stored lock addresses blocked on by other threads.
Similarly, these rogue unlock accesses may unblock a yield-looping thread.
These heuristics have enabled \landslide to test over 100 student thread libraries from the past 5 semesters fully automatically.

>>>>

\section{Logos}
% TODO camready
%\section{Landslide}
\label{sec:landslide}

@@

Conceptually,
\landslide is designed as follows.
Using several annotations in the kernel source, \landslide tracks
%the scheduler state to know
which threads are running or runnable.
(For testing thread libraries, these annotations are already provided in the reference kernel binary;
for testing kernels, students must add them by hand.)
When execution reaches certain interesting {\em preemption points},
\landslide triggers artificial clock interrupts
to force the scheduler to run a different thread.
When a test finishes execution according
to one pattern of thread switches,
\landslide rewinds the simulation state
and resumes the test according to a different thread interleaving.
After each instruction,
\landslide applies several bug-detection predicates to
the program state to detect
illegal heap accesses,
deadlock, infinite loops, and panics.

@@

{\bf State space exploration.}
To trigger the interleaving which exposes this bug, threads must be preempted when invoking the synchronization primitives (either when \x{cond_wait} blocks or at the end of \x{mutex_unlock}).
\landslide automatically considers these boundaries to be {\em preemption points}:
candidate sites for artificial timer-driven thread switches.
It can also add preemption points dynamically,
as identified by an integrated data-race analysis \cite{hybriddatarace,fasttrack},
to expose shared-memory bugs when synchronization is absent.

The branching possibilities arising from each preemption point produce an execution tree, or state space,
as shown in Figure \ref{fig:paradise}(b).
%As these state spaces are exponentially-sized,
\landslide uses Dynamic Partial Order Reduction (DPOR) \cite{dpor} to identify and prune equivalent interleavings
to mitigate the exponential explosion of these state spaces.
% In principle, the combination of synchronization API and data-race preemption points is sufficient to expose all bugs that are reachable by any combination of timer interrupts on any instructions.
This reduction requires only tracing the memory accesses of each thread,
so we are able to perform it behind-the-scenes, without consuming any student attention.

Apart from which test program to run, \landslide's user interface requires just one option:
students specify how long they are willing to wait.
%Considering this CPU budget,
% TODO CAMREADY: deanonymize
\landslide then uses state space estimation \cite{estimation-anonymized} to predict
% TODO CAMREADY: deanonymize
which combinations of preemption points will produce appropriately-sized state spaces for the given CPU-time budget \cite{quicksand-anonymized}.
If no bug is found before time runs out, \landslide displays a progress report summarizing the estimated remaining time for each state space,
so ambitious students can decide whether to increase the time limit and retry.

@@

When \landslide identifies a bug, it outputs an HTML-formatted
{\em preemption trace}.
Figure~\ref{fig:trace} shows a typical output for the example bug in Figure \ref{fig:paradise}.
The trace reports what kind of bug was detected
and each preemption point encountered in the current interleaving:
which thread was running, a trace of its stack when it was preempted, and the thread that ran next.
Hence, students can bypass tediously reconstructing the schedule using \x{printf},
and hopefully better understand the concurrent execution that exposed the bug,

@@

For the thread library project, \landslide instruments student implementations automatically.
We provide a custom reference kernel binary equipped with the kernel instrumentations (described below)
so \landslide can track threads' lifecycles. %the lifecycle of threads.
The userspace \x{malloc} and \x{assert} are also provided in the basecode, and tracked similarly.
\landslide's data-race analysis must understand the locking discipline,
so it searches the student's code for functions named \x{mutex_lock} and \x{mutex_unlock} (as mandated by the assignment handout),
and assumes they at least provide mutual exclusion\footnote{
	To relax this assumption, we provide a special test case for mutexes,
	in which \landslide will preempt and search for data races in the mutex implementation itself.
}.
All other thread library components are treated as unknown, and subject to preemption anywhere.

To test \pebbles kernels directly, they must be annotated to inform \landslide of certain important concurrency events.
Because the kernel project leaves internal code structure as an open-ended design decision,
\landslide cannot search for specific function names. % like it does for P2s.
Instead, we provide a set of functions named with the prefix
% TODO camready
\texttt{tell\_\landslide}
%\x{tell_landslide}
to denote when a thread runs \x{fork}, \x{sleep}, or \x{vanish}, when a thread is added to or removed from the runqueue, and when a thread becomes blocked on a mutex.
Users must also list all scheduler, context switch, and timer handling routines, so \landslide can ignore their memory accesses for the sake of DPOR
(otherwise, threads would always conflict with each other and DPOR would achieve no reduction).

@@

Student submissions may behave in extraordinary ways, violating the assumptions of many modern model checkers.
\landslide refuses to test code which fails the hurdle criteria (\sect{\ref{sec:grading}}),
but it must support testing any legal, if distasteful, solution.
%Hence, we added several heuristic
For example, students may use potentially infinite \x{yield} or \x{xchg} loops rather than the mutex abstraction. %(though it is discouraged).
To avoid getting stuck in a cyclic (infinite) state space, \landslide keeps a ``yield loop'' counter for each thread;
when a thread \x{yield}s 10 times without otherwise making progress, \landslide marks it blocked.
%
In extreme cases, this may erroneously seem like deadlock when all threads \x{yield} many times coincidentally.
To avoid wasting precious student attention with false alarms, % the ultimate sin
\landslide will reset all blocked threads' yield-loop counters,
retrying potentially up to 128 times before reporting a true deadlock.

Students may also break abstraction by open-coding a \x{mutex_unlock} sequence.
\landslide must track lock operations for accurate data-race analysis,
so to catch these patterns, it
%learns the size of the \x{mutex} struct, then
checks accesses against stored lock addresses blocked on by other threads.
Similarly, these rogue unlock accesses may unblock a yield-looping thread.

%So far,
These heuristics have enabled \landslide to test over 100 student thread libraries
%from the past 5 semesters
fully automatically
(47 in progress in our evaluation, plus
79 submitted over the past 5 semesters).

### 2016-08-27 sigcse-17/paper.tex
sigcse: rename double blind name of pebbles/landslide
sigcse: garth's revisions

% TODO camready
\newcommand\carnegiemellon{{\em Double-Blind University}\xspace}
\newcommand\cmu{DBU\xspace}
\newcommand\pebbles{\cmu-OS\xspace}
\newcommand\landslide{Landslide\xspace}
\newcommand\fourten{CS999\xspace}
\newcommand\twothirteen{CS888\xspace}
%\newcommand\carnegiemellon{{\em Carnegie Mellon University}\xspace}
%\newcommand\cmu{CMU\xspace}
%\newcommand\pebbles{Pebbles\xspace}
%\newcommand\landslide{Landslide\xspace}
%\newcommand\fourten{15-410}

@@

In \carnegiemellon's undergraduate operating systems course
students learn how to write correct and principled concurrent code through
a series of low-level programming projects.
The curriculum includes implementing hardware device drivers and a user-level
threading library, and culminates in a six-week long project in which students
build from the ground up a fully-preemptible, UNIX-like kernel that can run
on real hardware.

We aspire to expand the scope of these projects in response to
typical hardware platforms'
increasing complexity,
especially massive multi-core.
To that end, we wish to reduce the time students spend
debugging concurrency bugs.
We present Landslide, a
%systematic testing
stateless model checker which
offers a more principled approach than stress testing.
% edupar version
%We report on our preliminary experience working with students using Landslide to find previously-overlooked bugs in their own code, and discuss future methods for making Landslide more accessible to struggling students.
% sigcse version
We report on our experience giving students Landslide during the thread library
%and kernel projects.
project,
%Although kernel-mode Landslide requires manual instrumentation, testing thread libraries is fully automatic.
showing that it allows them to find and fix previously-overlooked bugs.
%% no :( :(
% Further, %even struggling students can take advantage of it,
% Landslide's users achieve higher concurrency scores on average in subsequent projects,
% suggesting that debugging with a model checker improves student retention of concurrency principles.

@@

%\section{Conclusions and Future Work}
\section{Future Work}

@@

% TODO camready
% \section{Acknowledgments}
% Many people have contributed to the development of
% 15-410 over the past decade.
% Stephen Muckle, currently at Qualcomm,
% led the switch to x86 projects running in Simics.
% Nathaniel Wesley Filardo,
% currently a Ph.D.\ candidate at Johns Hopkins,
% % increased the sanity of the build infrastructure,
% designed the \x{swexn()} system call and helped critique this paper.
% The current reference kernel, called ``pathos,''
% was written by Michael J. Sullivan,
% currently a CMU CS Ph.D.\ candidate,
% and Elly Fong-Jones, currently at Google.
% The SMP infrastructure code was contributed by
% Ryan Pearl, currently at Dropbox.
% Joshua Wise, currently at NVIDIA,
% led several kernel-extension projects.
% % including the hypervisor project.
% The course has benefitted
% greatly from free educational Simics licenses
% donated by Virtutech/Wind River.
% Code-size numbers were generated using David A.\ Wheeler's
% ``SLOCCount.''
% Work on Landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.

>>>>

% TODO camready
\newcommand\carnegiemellon{{\em Double-Blind University}\xspace}
\newcommand\cmu{DBU\xspace}
\newcommand\pebbles{EthOS\xspace}
\newcommand\landslide{Logos\xspace}
\newcommand\fourten{CS999\xspace}
\newcommand\twothirteen{CS888\xspace}
%\newcommand\carnegiemellon{{\em Carnegie Mellon University}\xspace}
%\newcommand\cmu{CMU\xspace}
%\newcommand\pebbles{Pebbles\xspace}
%\newcommand\landslide{Landslide\xspace}
%\newcommand\fourten{15-410}

@@

In \carnegiemellon's undergraduate operating systems course
students learn how to write correct and principled concurrent code through
a series of low-level programming projects.
The curriculum includes implementing hardware device drivers and a user-level
threading library, and culminates in a six-week long project in which students
build from the ground up a fully-preemptible, streamlined UNIX-like kernel that can run
on real hardware.

We aspire to expand the scope of these projects in response to
typical hardware platforms'
increasing complexity,
especially massive multi-core.
To that end, we wish to reduce the time students spend
debugging concurrency bugs.
We present \landslide, a
%systematic testing
stateless model checker which
offers a more principled approach than stress testing.
% edupar version
%We report on our preliminary experience working with students using \landslide to find previously-overlooked bugs in their own code, and discuss future methods for making \landslide more accessible to struggling students.
% sigcse version
We report on our experience giving students \landslide during the thread library
%and kernel projects.
project,
%Although kernel-mode \landslide requires manual instrumentation, testing thread libraries is fully automatic.
showing that it allows them to find and fix previously-overlooked bugs.
%% no :( :(
% Further, %even struggling students can take advantage of it,
% \landslide's users achieve higher concurrency scores on average in subsequent projects,
% suggesting that debugging with a model checker improves student retention of concurrency principles.

@@

\section{Conclusions and Future Work}

@@

% TODO camready
% \section{Acknowledgments}
% Many people have contributed to the development of
% 15-410 over the past decade.
% Stephen Muckle, currently at Qualcomm,
% led the switch to x86 projects running in Simics.
% Nathaniel Wesley Filardo,
% currently a Ph.D.\ candidate at Johns Hopkins,
% % increased the sanity of the build infrastructure,
% designed the \x{swexn()} system call and helped critique this paper.
% The current reference kernel, called ``pathos,''
% was written by Michael J. Sullivan,
% currently a CMU CS Ph.D.\ candidate,
% and Elly Fong-Jones, currently at Google.
% The SMP infrastructure code was contributed by
% Ryan Pearl, currently at Dropbox.
% Joshua Wise, currently at NVIDIA,
% led several kernel-extension projects.
% % including the hypervisor project.
% The course has benefitted
% greatly from free educational Simics licenses
% donated by Virtutech/Wind River.
% Code-size numbers were generated using David A.\ Wheeler's
% ``SLOCCount.''
% Work on \landslide was sponsored by the U.S. Army Research Office under grant number W911NF0910273.
